Measuring causal influence with\\ back-to-back regression: the linear case - supplementary material

### 2020-03-04 paper/introduction.tex
up

First, causal and non-causal factors may be numerous and collinear.
% In physics,
% for example, one may be challenged to identify whether fusion is caused by a
% change in temperature or a change in pressure, as these two factors may, at
% first, be difficult to manipulate independently. This issue becomes increasingly
% pronounced as the number of potential factors increases.
In neuroscience, for
example, it can be challenging to identify whether word frequency modulates
brain activity during reading. Indeed, the
frequency of words in natural language covaries with other factors such as their
length (short words are more frequent than long words) and their categories
(determinants are more frequent than adverbs)
\citep{kutas2011thirty,pegado2014timing}. Instead of selecting a set of words
that controls for all of these factors simultaneously, it is thus common to use
a \emph{forward} "encoding model", i.e. to fit a linear regression to predict observations
(e.g. brain activity) from a minimal combination of competing factors (e.g.
word length, word frequency), and analytically investigate,
the estimated contribution of each factor from the model's coefficients
\citep{friston1994statistical,naselaris2011encoding,weichwald2015causal,
king2018encoding,huth2016natural}.

@@

Both \emph{forward} and \emph{backward} modeling have competing benefits and drawbacks.
Specifically, forward modeling disentangles the independent contribution of
collinear factors, but does not combine multidimensional observations. By
contrast, backward modeling combines multiple observations, but does not
disentangle collinear factors \citep{weichwald2015causal,
hebart2018deconstructing, king2018encoding}. To combine some of the benefits of forward
and backward modeling, several authors have proposed to use cross-decomposition
techniques such as Partial Least Squares (PLS) and Canonical Correlation
Analysis (CCA) \citep{de2019multiway}. CCA and PLS aim to find, from two sets of
data $X$ and $Y$, the components $H$ and $G$ were $XH$ and $YG$ are maximally
correlated or maximally covarying respectively. Because CCA and PLS are based on
a generalized eigen decomposition, their resulting coefficients are mixing the
features of $X$ and $Y$ in a way that makes them notoriously difficult to
interpret \citep{lebart1995statistique}.

@@

After detailing B2B method and proving its convergence
(Section~\ref{sec:theorem}), we show with synthetic data that it outperforms
state-of-the-art forward, backward and cross-decomposition techniques in
disentangling causal factors (Section~\ref{sec:experiment_synthetic}). Finally,
we apply B2B to a large neuroimaging dataset and reveal that distinct but
collinear word features lead to distinguishable brain representations
(Section~\ref{sec:experiment_real}).

>>>>

First, causal and non-causal factors may be numerous and partially correlated.
% In physics,
% for example, one may be challenged to identify whether fusion is caused by a
% change in temperature or a change in pressure, as these two factors may, at
% first, be difficult to manipulate independently. This issue becomes increasingly
% pronounced as the number of potential factors increases.
In neuroscience, for
example, it can be challenging to identify whether word frequency modulates
brain activity during reading. Indeed, the
frequency of words in natural language covaries with other factors such as their
length (short words are more frequent than long words) and their categories
(determinants are more frequent than adverbs)
\citep{kutas2011thirty,pegado2014timing}. Instead of selecting a set of words
that controls for all of these factors simultaneously, it is thus common to use
a \emph{forward} "encoding model", i.e. to fit a linear regression to predict observations
(e.g. brain activity) from a minimal combination of competing factors (e.g.
word length, word frequency), and analytically investigate,
the estimated contribution of each factor from the model's coefficients
\citep{friston1994statistical,naselaris2011encoding,weichwald2015causal,
king2018encoding,huth2016natural}.

@@

Both \emph{forward} and \emph{backward} modeling have competing benefits and drawbacks.
Specifically, forward modeling disentangles the independent contribution of
correlated factors, but does not combine multidimensional observations. By
contrast, backward modeling combines multiple observations, but does not
disentangle factors that are linearly correlated \citep{weichwald2015causal,
hebart2018deconstructing, king2018encoding}. To combine some of the benefits of forward
and backward modeling, several authors have proposed to use cross-decomposition
techniques such as Partial Least Squares (PLS) and Canonical Correlation
Analysis (CCA) \citep{de2019multiway}. CCA and PLS aim to find, from two sets of
data $X$ and $Y$, the components $H$ and $G$ were $XH$ and $YG$ are maximally
correlated or maximally covarying respectively. Because CCA and PLS are based on
a generalized eigen decomposition, their resulting coefficients are mixing the
features of $X$ and $Y$ in a way that makes them notoriously difficult to
interpret \citep{lebart1995statistique}.

@@

After detailing B2B method and proving its convergence
(Section~\ref{sec:theorem}), we show with synthetic data that it outperforms
state-of-the-art forward, backward and cross-decomposition techniques in
disentangling causal factors (Section~\ref{sec:experiment_synthetic}). Finally,
we apply B2B to a large neuroimaging dataset and reveal that distinct but
linearly-correlated word features lead to distinguishable brain representations
(Section~\ref{sec:experiment_real}).

### 2020-03-04 paper/meg_exps.tex
up

We aim to identify the word features that cause a variation in brain responses.
We consider four distinct but collinear features.
%
First, 'Word Length' refers to the total number of letters. Word Length is
expected to specifically cause a variation in the early evoked MEG responses
(i.e. from 100 ms after stimulus onset) elicited by the retinotopically-tuned
visual cortices (e.g. \citep{pegado2014timing}.).
%
Second, 'Word Frequency' indexes how frequently each word appears in Dutch and
was derived with the the Zipf logarithmic scale of \citep{van2014subtlex}
provided by the WordFreq package \citep{speerwordfreq}. Word Frequency is
expected to specifically cause a variation in the late evoked MEG responses
(i.e. from 400 ms), because it variably engages semantic processes in the
temporal cortices \citep{kutas2011thirty}.
%
Third, 'Word Function' indicates whether each word is a content word (i.e. a
noun, a verb, an adjective or an adverb) or a function word (i.e. a preposition,
a conjunction, a determinant, a pronoun or a numeral), and was derived from
Spacy's part of speech tagger \citep{spacy2}. To our knowledge, this feature has
not been thoroughly investigated with MEG. Its causal contribution to reading
processes in the brain thus remains unclear.
%
Finally, to verify that B2B and other methods would not inadequately identify
non-causal features, we added a dummy feature, constructed from a noisy
combination of Word Length and Word Frequency: $dummy = z(length) + z(frequency)
+ \mathcal{N}$, where $z$ normalizes features and $\mathcal{N}$ is a random
vector sampling Gaussian distribution (all terms thus have a zero-mean and a
unit-variance). This procedure yields an $X \in \mathbb{R}^{n \times d_x}$
matrix of $n\approx$ 2,700 words by $d_x=4$ features for each subject. Each of
the columns of $X$ is normalized to have a mean and a standard deviation of 0
and 1 respectively.

@@

\subsubsection{Results}
We compared the ability of Forward regression, Backward regression, CCA, PLS and
B2B to estimate the causal contribution of four distinct but collinear features
on brain evoked responses to words.

@@

As expected, the Backward model reveals a similar decoding time course for Word
Length and Word Frequency, even though these features are known to specifically
influence early and late MEG responses respectively \citep{kutas2011thirty}. In
addition, the same decoding time course was observed for the dummy variable.
These results illustrate that backward modeling cannot be used to estimate the
causal contribution of collinear features.

>>>>

We aim to identify the word features that cause a variation in brain responses.
We consider four distinct but linearly-correlated features.
%
First, 'Word Length' refers to the total number of letters. Word Length is
expected to specifically cause a variation in the early evoked MEG responses
(i.e. from 100 ms after stimulus onset) elicited by the retinotopically-tuned
visual cortices (e.g. \citep{pegado2014timing}.).
%
Second, 'Word Frequency' indexes how frequently each word appears in Dutch and
was derived with the the Zipf logarithmic scale of \citep{van2014subtlex}
provided by the WordFreq package \citep{speerwordfreq}. Word Frequency is
expected to specifically cause a variation in the late evoked MEG responses
(i.e. from 400 ms), because it variably engages semantic processes in the
temporal cortices \citep{kutas2011thirty}.
%
Third, 'Word Function' indicates whether each word is a content word (i.e. a
noun, a verb, an adjective or an adverb) or a function word (i.e. a preposition,
a conjunction, a determinant, a pronoun or a numeral), and was derived from
Spacy's part of speech tagger \citep{spacy2}. To our knowledge, this feature has
not been thoroughly investigated with MEG. Its causal contribution to reading
processes in the brain thus remains unclear.
%
Finally, to verify that B2B and other methods would not inadequately identify
non-causal features, we added a dummy feature, constructed from a noisy
combination of Word Length and Word Frequency: $dummy = z(length) + z(frequency)
+ \mathcal{N}$, where $z$ normalizes features and $\mathcal{N}$ is a random
vector sampling Gaussian distribution (all terms thus have a zero-mean and a
unit-variance). This procedure yields an $X \in \mathbb{R}^{n \times d_x}$
matrix of $n\approx$ 2,700 words by $d_x=4$ features for each subject. Each of
the columns of $X$ is normalized to have a mean and a standard deviation of 0
and 1 respectively.

@@

\subsubsection{Results}
We compared the ability of Forward regression, Backward regression, CCA, PLS and
B2B to estimate the causal contribution of four distinct but linearly-correlated features
on brain evoked responses to words.

@@

As expected, the Backward model reveals a similar decoding time course for Word
Length and Word Frequency, even though these features are known to specifically
influence early and late MEG responses respectively \citep{kutas2011thirty}. In
addition, the same decoding time course was observed for the dummy variable.
These results illustrate that backward modeling cannot be used to estimate the
causal contribution of correlated features.

### 2020-03-04 paper/paper.tex
up

\usepackage{lineno,hyperref}
\modulolinenumbers[5]

@@

\title{Discriminating the Linear Influences of Collinear Factors \\from Multivariate Observations: the Back-to-Back Regression}

@@

\begin{abstract}
Identifying causes solely from observations can be particularly challenging when
i) potential factors are difficult to manipulate individually and ii)
observations are multi-dimensional. To address this issue, we
introduce ``Back-to-Back'' regression (B2B), a linear method designed to efficiently
measure, from a set of collinear factors, those that most plausibly account for
multidimensional observations. First, we prove the consistency of B2B, its links to
other linear approaches, and show how it provides a robust, unbiased and
interpretable scalar estimate for each factor.
Second, we use a variety of simulated data to show that B2B outperforms
least-squares regression and cross-decomposition techniques (e.g. canonical
correlation analysis and partial least squares) on causal identification when
both the factors and the observations are collinear.
Finally, we apply B2B to magneto-encephalography of 102 subjects
recorded during a reading task to test whether our method appropriately disentangles
the respective contribution of word length and word frequency - two collinear
factors known to cause early and late brain responses respectively. The results
show that these two factors are better disentangled with B2B than with
other standard techniques. We finally discuss the limits of B2B.
\end{abstract}

@@

Forward and cross-decomposition models have been used to identify the causal
contribution of collinear features onto multi-dimensional observations (e.g.
\citep{naselaris2011encoding}). These approaches typically lead to multiple
coefficients for each features (i.e. one per dimension of $Y$ or one per
component respectively). Furthermore, these coefficients can be difficult to
summarize into a single causal estimate. By contrast, B2B quickly (Fig.
\ref{fig:duration}) leads to a single unbiased scalar values $\hat E$ tending
towards 1 and 0 for causal and non-causal features respectively.

@@

\clearpage
\newpage

@@

\section{Appendices}
\input{appendix}

>>>>

\usepackage{lineno,hyperref}
% \modulolinenumbers[5]

@@

\title{Discriminating the Influence of Correlated Factors \\from Multivariate Observations: the Back-to-Back Regression}

@@

\begin{abstract}
Identifying causes solely from observations can be particularly challenging when
i) potential factors are difficult to manipulate independently and ii)
observations are multi-dimensional. To address this issue, we
introduce ``Back-to-Back'' regression (B2B), a linear method designed to efficiently
measure, from a set of correlated factors, those that most plausibly account for
multidimensional observations. First, we prove the consistency of B2B, its links to
other linear approaches, and show how it provides a robust, unbiased and
interpretable scalar estimate for each factor.
Second, we use a variety of simulated data to show that B2B outperforms
least-squares regression and cross-decomposition techniques (e.g. canonical
correlation analysis and partial least squares) on causal identification when
the factors and the observations are partially collinear.
Finally, we apply B2B to magneto-encephalography of 102 subjects
recorded during a reading task to test whether our method appropriately disentangles
the respective contribution of word length and word frequency - two correlated
factors known to cause early and late brain responses respectively. The results
show that these two factors are better disentangled with B2B than with
other standard techniques.
\end{abstract}

@@

Forward and cross-decomposition models have been used to identify the causal
contribution of correlated factors onto multi-dimensional observations (e.g.
\citep{naselaris2011encoding}). These approaches typically lead to multiple
coefficients for each features (i.e. one per dimension of $Y$ or one per
component respectively). Furthermore, these coefficients can be difficult to
summarize into a single causal estimate. By contrast, B2B quickly (Fig.
\ref{fig:duration}) leads to a single unbiased scalar values $\hat E$ tending
towards 1 and 0 for causal and non-causal features respectively.

@@

\section{Acknowledgements}
We are thankful to Gael Varoquaux and Alexandre Gramfort for the valuable
feedback. This work was supported by ANR-17-EURE-0017 and the Fyssen Foundation
to JRK.

@@

\newpage
\clearpage
\section{Appendices}
\input{appendix}

### 2020-03-04 paper/supplementary.tex
supp
up

     Consider the B2B model from Equation $$Y = (XE + N)F$$ $N$ centred and full rank noise.

     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimise

@@

\section{Additional Figures}

\begin{figure}[t!]
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_supp.pdf}
  \caption{}
  \label{fig:meg_supp}
\end{figure}

>>>>

     Consider the B2B model from Equation $$Y = (XE + N)F$$ $N$ centered and full rank noise.

     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimize

@@
